%%  By Himarsha R. Jayanetti and Dr. Michele C. Weigle
%%  April 16, 2024

%% The font size of the dissertation should be consistent at 12 pt
\documentclass[12pt]{report}

%%%%% Thesis or Dissertation Selection %%%%% 
\usepackage[diss]{odusci}  
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
% for MS thesis, replace diss with thesis
% \usepackage[thesis]{odusci}

%%%%% Citation Style %%%%% 
%% Uncomment the appropriate line for your discipline
%% IEEE
%% Using backend=bibtex for environments where biber is unavailable.
\usepackage[backend=bibtex, style=ieee, sorting=nyt, citestyle=numeric-comp, maxnames=10, minnames=10]{biblatex}
%% APA
% \usepackage[backend=bibtex, style=apa, sorting=nyt, maxnames=10, minnames=10]{biblatex}
%% Science 
% \usepackage[backend=bibtex, style=science, sorting=nyt, maxnames=10, minnames=10]{biblatex}

%% add BibTeX file
\addbibresource{ref.bib}

% Adjust the options below to refine the appearance of code listings
\lstset{
  basicstyle=\sffamily,
  columns=fullflexible,
  frame=single,
  breaklines=true,
  moredelim=**[is][\color{red}]{@}{@},
  escapechar=\%
}

%% ***DO NOT CHANGE OR COMMENT THE NEXT LINE*** 
\setlength\intextsep{1.69cm} % to set space above/below the floats and text 
\begin{document}

%%%%% TITLE PAGE %%%%% 
\title{INTERPRETABLE SPARSE MODELING OF LONGITUDINAL SIGNALS VIA CRITICAL-RANGE RECTIFICATION AND ANYTIME RULE COMPRESSION}

\author{Jason Orender}

%% US degree
\degrees{B.S. December 1993, University of Texas at Austin\\
		 M.S. December 2003, George Mason University\\
         M.S. May 2018, Old Dominion University}

%% Department
\dept{Computer Science}

%% Generates Copyright page, do not uncomment for thesis or dissertation
%\copyrightfalse

%% submit date must use either "May", "August", or "December" for the month
\submitdate{August 2026}
% \submitdate{December 2024}

%% Default is for a single adviser
\principaladviser{Jiangwen Sun}
\member{Mohammed Zubair}
\member{Yaohang Li}
\member{Yet Nguyen}

%% Co-advisers - uncomment below if required
%\coadvisers{{Jiangwen Sun}
%\coadvisers{Mohammed Zubair}
%\member{Yaohang Li}
%\member{Yet Nguyen}
%% See https://github.com/oduwsdl/odusci-etd-template/blob/main/README.md
%% for instructions on adjusting odusci.sty to allow co-advisers

%%%%% FRONT MATTER %%%%% 

%% Abstract - maximum 350 words
\abstract{High-dimensional longitudinal data are common in clinical monitoring, industrial control systems, and other sensor-driven domains. When outcomes are governed by go/no-go thresholds and delayed effects, standard sparse models such as L1-regularized logistic regression often become unstable after lag expansion because multicollinearity undermines consistent support recovery. At the same time, many stakeholders require models that are interpretable as simple rules and that can be trained and evaluated under constrained compute.
This prospectus proposes an end-to-end framework that addresses these constraints using three stages: (i) critical-range rectification, which converts each lagged continuous measurement into a \{-1,+1\} indicator denoting whether it lies within a data-driven range observed during events; (ii) sparse model fitting using an L1-regularized logistic baseline on the rectified design; and (iii) anytime rule compression (“logic polishing”), which collapses small-magnitude coefficients to produce a compact m-of-K rule model tuned to maximize Youden’s J at a chosen operating point. Preliminary studies on synthetic data with known ground-truth rules and on real-world datasets (including the HAI industrial control system) suggest that rectification improves feature-and-lag attribution, reduces false positives, and can decrease end-to-end training time.
A tractable proof-of-concept theoretical analysis uses the zero-threshold arcsin relation under joint normality to show generally how sign binarization contracts pairwise correlations, improves conditioning of the Gram matrix, and increases the likelihood of satisfying the LASSO irrepresentable condition. The proposed dissertation work will (a) validate the proof-of-concept assertions, (b) quantify the contribution of each pipeline stage via ablations and stability studies, and (c) expand evaluation to additional longitudinal domains and baselines.} 

%% Dedication - optional, 1 page max
\dedication{I dedicate my dissertation to my wife Danielle.}

%% Acknowledgements
\acknowledge{I would like to thank my committee, and especially Drs. Sun and Zubair, without whose guidiance this document and the work it represents would not have been possible. 
%% If you want multiple paragraphs, must add a newline and indent as below
%\\ \indent 
%Morbi vulputate feugiat diam, rhoncus vehicula eros posuere vitae. Pellentesque sed egestas metus. Vestibulum libero elit, euismod quis vestibulum nec, vulputate at risus. In accumsan sem eu porta ultrices. Quisque sagittis massa felis, eget varius metus semper id. Vivamus tincidunt ipsum sem, sit amet eleifend diam tempor nec. Vivamus elementum, diam id lobortis mattis, nisi lacus.
}

%% Nomenclature - optional, uncomment below if required 
% \nomenclature{
% \begin{tabular}{ll}
% \textit{A} & Amplitude Ratio, (No Unites) \\   
% \textit{C} & Centroid of pipe, inches \\ 
% \textit{Do} & Outside Diameter of Pipe, inches \\ 
% \textit{ZNom} & Section Modulus, in3 \\ 
%  \end{tabular}
% }
%% See https://github.com/oduwsdl/odusci-etd-template/blob/main/README.md
%% for instructions on adjusting odusci.sty to include nominclature

%%%%% Before and After Acknowledgement Page %%%%%
%% ***DO NOT CHANGE OR COMMENT THE NEXT TWO LINES*** 
    \beforepreface
    \afterpreface

%%%%% CHAPTERS %%%%% 
%% It is recommended to put the contents of each chapter in a 
%% separate .tex file in the "Chapters" folder and include them as below.  
\input{Chapters/01_introduction}
\input{Chapters/02_background}
\input{Chapters/03_relatedwork}
\input{Chapters/04_resquestions}
\input{Chapters/05_rq1}
\input{Chapters/06_rq2}
\input{Chapters/07_rq3}
\input{Chapters/88_futurework}
\input{Chapters/89_conclusion}

\cleardoublepage

%%%%% REFERENCES %%%%% 
\addtocontents{toc}{\vspace*{12pt}} % gap between last chapter and references in the table of contents
\printbibliography[heading=bibintoc, title={REFERENCES}]

%%%%% APPENDICES %%%%% 
%% Comment the lines below if you do not have any appendices
\begin{Appendices}
\include{Chapters/98_appendices} % First appendix file
% \include{Chapters/98_appendices2} % Second appendix file
% You can add more .tex files as needed
\end{Appendices}

%%%%% VITA - max 1 page %%%%% 
\include{Chapters/99_vita}

\end{document}
